---
title: "Working with your dinosauR"
author: "The R4All Team"
date: "13/03/2017"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

_'We will one day all be dinosauRs'_

Fashion happens, even in R.  It seems like everyone HAS to learn the tidyverse, and at least the data management tools of dplyr and the graphing tools of ggplot2.  It's so hot!  It is quite cool. But it has created dinosauRs - people weaned and reared on the milk of base R.  base R is nutritious and good.  But it scares the new tidypeople.  And the base R dinosauRs are sometimes scared or even ambivalent to tidypeople.   

Why change? It's consistent, fast, effective and makes working with data and graphs for data wrangling, data science and plain old science more easy and more fun.  We think. But it is not required, and not the end of the world if someone doesn't know or use it.

And we are acutely aware that in moving our teaching material and our book, _Getting Started with R: An Introduction for Biologist_ to the tidyverse, and particularly to ggplot2 and dplyr for graphing and data manipulation, we have created a situation where students may be learning how to do things differently than their supervisors, fellow students, post-docs and people/help groups they interact with online for help.  We are acutely aware, because it's happened around us, with our own students and colleagues.  This isn't bad.  It is fun.  Really.

But, getting to the point where you do  statistics - e.g. organising your data and making a good figure before you start analysing the data - well, this has changed for tidypeople.  This reflects the reality is that there are just now more ways to do things than before. One way or another, we'll become dinosauRs.

This blog post is designed to highlight some of what changed. It's also designed to help you in 'working with dinosauRs'.  What follows is a set of examples of doing things in base R and dplyr or ggplot2.  It is not exhaustive.  Just a primer for liaising with dinosauRs.


# Two ways to import data

First we load some packages and import the data. But we import the data each of two ways, the base way and the tidyverse way.


```{r, message = FALSE}
# libraries you need
library(dplyr)
library(ggplot2)
library(readr)

# get the data into R with the base R function
compensation_br <- read.csv('compensation.csv')

## get the data into R with the tidyverse function
compensation_tv <- read_csv('compensation.csv')

```

(In case you didn't spot the difference, its the full stop or underscore between `read`and `csv`.)

What's the difference here? One big one is `read.csv`  converts character variables to factors. Many people prefer this doesn't happen automatically (`read_csv` doesn't do this). Another is `read.csv` creates a `data.frame` while `read_csv`creates a `tibble`. One nice thing about a `tibble` is how it looks in the Console -- a much more considerate view than one often get with a big dataframe:

```{r}
compensation_tv
```



# Messing with your data: dplyr vs. base

In our book, we provide a table with comparative code from dplyr and base R.  Here we expand a bit on this, providing a New R User and new tidyperson template for speaking and interacting with your pet dinosauR.

```{r, out.width = "800px", echo = FALSE}
knitr::include_graphics('old_new_comparison_table.pdf')
```

dplyr is a toolbox with a set of functions, each optimised to do one thing and one thing very well, and very fast. It is also designed to help you by first giving the data frame (or more likely [tibble](https://cran.r-project.org/web/packages/tibble/vignettes/tibble.html)) in which all of your variables are stored, and then asking for things to happen.  This is a highly standardised toolbox.  We like this.

Lets work with three examples  - getting a column, getting rows and subsetting your data based on values of a variable. We look at base and dplyr methods to do each.  

To run the code, you'll need to download the dataset `compensation.csv` from the R4All website (get the zip file of all the datasets, upack it, and you'll see the `compensation.csv` dataset.



## Get columns with dplyr

With dplyr, we can select a column with one function two ways:

```{r}
# use select().  The first argument is the data frame and the second is the column.
select(compensation_tv, Fruit)

# or
# select(compensation_tv, 2) # gets the second column, which is Fruit!
```

The important detail here is that `dplyr` functions alway take the data frame as the first argument and ALWAYS return a `tibble`.

## Get columns with base R

How can you do this in base R?  What might you see from the dinosauRs?  Well, lots of different ways to do the 'same' thing.... almost.  Lets look at 4 ways, and see what happens.

```{r}
# Base Method 1: $
compensation_br$Fruit

# Base Method 2: [, 'column name' ]
compensation_br[,'Fruit']

# Base Method 3: [, column_number]
compensation_br[,2]

# Base Method 4: subset()
subset(compensation_br, select = Fruit)
```

What do we notice?  Well, not only are there 4 ways to do the same thing, only one of them actually returns 'exactly' what `dplyr`'s select() does, a data frame of Fruit.  Which one?  `subset()`.

What are the other three returning?  They are returning vectors... e.g. just the numbers. They are not organised into a single column data frame with a column lable called Fruit.  This might be what you want sometimes... it might not other times.  Keep in mind that this, we believe, is one of the fun bits of R.



## Get a row (or some rows) with dplyr

Selecting rows in R is also as variable as selecting columns.

Lets recall the dplyr method - `slice()` or `filter()`

```{r}
slice(compensation_tv, 2) # get the second row
```

Notice that again, dplyr function `slice()` starts with the data frame as it's first argument, and then specifies the row number.  And importantly, we now have a 1 row, 3 column data frame.

Lets recall how dplyr does this using `filter()`

```{r}
filter(compensation_tv, Fruit > 80)
```


## Get a row (or some rows) with base R

Base R provides one way to do this.

```{r}
# Base Method 1: [ ]
compensation_br[2,]
```

Note where the comma is, and that the outcome is the same (it's got the right numbers) AND it has returned a data frame.  Interesting.... eh?


The final bit of base-dplyr punch up we want to compare involves subsetting your data.  Lets imaging we want to return all rows that have Fruit values > 80 kg.  Lets work through several ways to do this with base R, and then shift to dplyr.

```{r}
# Base Method 1: [ ]
compensation_br[compensation_br$Fruit>80,]

#Base Method 2: subset()
subset(compensation_br, subset = Fruit > 80)
```


#### Getting some rows and some coluns... the grand challenge

However, lets try and make this a bit more complicated to see how different base methods stack up to dplyr tools.  Lets now not only request `Fruit > 80`, but also JUST return the Root values where `Fruit > 80`.

```{r}
# Base Method 1: [ ] + $
compensation_br[compensation_br$Fruit>80,]$Root

# Base Method 2: subset() + $
subset(compensation_br, subset = Fruit > 80)$Root

#Base Method 2: subset()
subset(compensation_br, subset = Fruit > 80, select = Root)
```

Notice three things here.  First, Base Method 1 involves `[ ]` and `$`.  It also now returns a vector, not a data frame.  Second, Base Method 2, using `subset()` and $ also returns a vector.  `subset()` actually returns a data frame, but $ returns a vector.  Base Method 3 uses only `subset()` and returns a data frame and reveals that the function `subset()` has two arguments that are indeed quite handy: `susbet =` and `select =`.  Very nice indeed.

What do we need to do to replicate the functionality of subset() in dplyr?  Given that dplyr functions tend to only do one thing, we actually need both the `filter()` and the `select()` function.  And, actually, there are two ways to do this: we call them the nested method and the piping method.

The nested method follows old skool ideas in R.  Functions within functions are evaluated from the inside out.

```{r}
select(filter(compensation_tv, Fruit>80), Root) # nested method
```

Here, `filter()` is first evaluated, producing a data frame, followed by applying to this new dataframe, on-the-fly, the `select()` function.

However, dplyr ships with another package built in... `magrittr`.  It's quite awesome.

magritter has a function `%>%` that is called the pipe.  It passes the stuff-on-the-left to the thing-on-the-right.  It works by starting with the data frame, and passing it on to a function that does stuff, and passing this stuff on to other functions that do stuff.... like this.

```{r}
compensation_tv %>% 
  filter(Fruit>80) %>% 
    select(Root) # piping method
```

Nice!  This captures the order in which you want things to happen, without thinking from the inside out.  Neither way of thinking is right or wrong, we note.  And with the dplyr functions, you always always deliver a data frame and get a data frame.


## Summary

Base R and dplyr ship with tools that perform very similar functions  The thing to remember with R and with your dinosaurRs, is that they are all good.  Just remember that dplyr is way more consistent in what it takes - always a data frame - and returns - always a data frame.  dplyr functions always, also, do one thing.  For example, `subset()` from base R can do two things... select and filter.  

Just remember to think about what you want before you start!





# Messing with your graphs: ggplot2 vs. base (vs. lattice)

We might as well add a little bit of base graphics versus ggplot2 for the fun of it.  

We'll make a 2 colour graph that corresponds to an ANCOVA model scenario - e.g. a continuous and a categorical explanatory/independent variable, and a numerical response/dependent variable.  We'll use the same data.

## Some base graphics

If you want to see more of these base graphics, feel free to consult the web, or our 1st edition book.

We adhere to the principle, in base graphics, that the question you are asking and the data you are plotting go hand-in-hand.  With the data we are using lets use Fruit as our response variable and Root as the continuous explanatory variable, and Grazing as our categorical explanatory variable.  Lets assume that we are looking at Fruit set on apple trees as a function of the grafted root stock width and whether cows have been allowed to graze underneath them or not (reducing competition by eating the grass, of course).

The `plot()` function that is at the heart of base graphics can take a formulae as it's first argument.  We always like(d) this.  We start with the _regression_ part of the model - `Fruit ~ Root` where `~` translates "is a function of...". Lets also not forget that `plot()` takes a data argument too.

```{r}
plot(Fruit ~ Root, data = compensation_br)
```

At this point, and to keep this focused, lets decide on a few things to do to this graph:

1)Let's colour the points by the Grazing factor
2)Let's double the size of the points, and make them filled
3)Lets customise the x- and y-axis labels, changing the words
4)Add a legend in the top left

```{r}
plot(Fruit ~ Root, data = compensation_br, 
    # double the font size with cex (character expansion
    # make the points solid (filled) using pch (point character)
    cex = 2, pch = 19,
    # colour the points by Grazing
    # brown for grazed, green for ungrazed (alphabetical rules)
    # this takes a vector of colours and allocates them alphabetically
    # to each level of granzing (Grazed and Ungrazed)
    col = c('brown','green')[Grazing],
    # x- and y-axis labels
    xlab = "Root Stock (mm)", ylab="Fruit Harvest (kg)"    
)

# add the legend, after the graph is made.
legend('topleft', col=c('brown','green'), 
       pch = 19, legend = c("Grazed","Ungrazed"))
```

## The ggplot2 hipsteR might do something like this.
Doing most of this in ggplot2 is easy, and some would suggest easier.  There are some things that are easier, and some that simply require knowing that base and ggplot2 graphics each have their own vocabulary.  Lets see what we can do.

We begin with the foundation layer of the graph.  Just the points, axes and labels.

```{r}
ggplot(compensation_tv, aes(x = Root, y = Fruit))+
  geom_point()
```

If you are not super up2speed with ggplot2:

1) it always starts with the data frame... that is the first argument
2) `aes()` are the aesthetics, and map the data (x and y variables, Root and Fruit) to the graph coordinate system.
3) `geom_point()` adds the points, at the right x-y locations (the coordinate system), using a point.

Lets customise it in the same way.  We'll come back to explain some of the syntax...

1) Let's colour the points by the Grazing factor
2) Let's double the size of the points, and make them filled
3) Lets customise the x- and y-axis labels, changing the words
4) Add a legend.... (!)

```{r}
ggplot(compensation_tv, aes(x = Root, y = Fruit, colour = Grazing))+
  geom_point(size = 2)+
  xlab("Root Stock (mm)") + ylab("Fruit Harvest (kg)")+
  theme_bw()
```

We've deliberately not finished, just to show what a few keywords in ggplot2 do.

1) colour, in the aesthetics, does what `col = c('brown','green')[Grazing]` does in base graphics, but uses some default colours.
2) `geom_point()` adds the points.  `size=2` in the `geom_point()` makes ALL points double size.
3) `theme_bw()` has removed the grey background, but left the gridlines.  There are packages with tonnes and tons of theme options, including one that can look JUST LIKE BASE! (ggthemes)

Note too that by specifying `colour()` in the aesthetics, a legend is produced and placed for you.

All of this is handy dandy.  However, people ALWAYS wonder how to change the colours.  They always wonder this with base graphics too.  It may be your dinosauR knows how to do this our way above, or another way.  There are many.

In ggplot2, for this kind of graph, there is a good way to do this.

```{r}
ggplot(compensation_tv, aes(x = Root, y = Fruit, colour = Grazing))+
  geom_point(size = 2)+
  xlab("Root Stock (mm)") + ylab("Fruit Harvest (kg)")+
  scale_colour_manual(values = c(Grazed = 'brown', Ungrazed = 'green'))+
  theme_bw()
```

## Summary

There you have it.  A quick introduction to the dinosauR you may be working with or have to collaborate with.  Don't get angry at the dinosauR.  base R is awesome and does a great many things a great deal goodly.  Most of the time, base v. tidyverse is not a battle at all. But... this is not only a tutorial to handle the dinosauR, but something the dinosauR can use to start gettin' hip with the youngsteRs.

# But... respect your DinosauRs

No disaster has wiped out the DinosauRs, or wiped out how they use R. They will still be very effective. Some of the DinosauRs will be very impressive animals. So don't write them off, and realise that there will likely be a lot to learn from them. Respect your dinosauRs! 
